The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-24 paper/ms.tex
mostly superficial revisions to 2.2. Added a reference to a paper supporting the assumption of a flat rotation curve near the solar circle.

\subsection{Epicyclic Approximation} \label{ssec:epi_action}
Before turning to numerical methods, we first write down analytically the
induced action error from either a midplane offset or a Galactocentric radius
offset. We make the epicyclic approximation, which assumes that the motion in
the $z$ and $R$ components of the orbit are decoupled and follow a simple
harmonic oscillator. This approximation should be an excellent description of
thin disk-like orbits and a good description of thick disk-like orbits. We
also make the assumption of a perfectly flat circular velocity curve with
velocity $v_c$.

Under this approximation, we can write down the vertical and radial components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
z(t) &= A_z \sin{(\kappa t + \alpha)}\\
R(t) &= A_R \sin{(\nu t + \beta)}\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the vertical and radial (or epicyclic)
frequencies, $A_z$ and $A_R$ are the amplitudes of the stellar motion in the
vertical and radial coordinates, and $\alpha$ and $\beta$ are arbitrary
constants. Similarly, the velocities of the orbit are given by,
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_z(t) &= A_z \kappa \cos{(\kappa t + \alpha)}\\
v_R(t) &= A_R \nu \cos{(\nu t + \beta)}\text{.}
\end{split}
\eeq

We can write down the actions in this approximation as
\citep[Section~3.5.3b][]{2008gady.book.....B},
\beq\label{eq:actions_epi}
\begin{split}
J_{\phi} &= R_g v_c \\
J_R &= \kappa A_R^2 \\
J_z &= \nu A_z^2\text{,}
\end{split}
\eeq
where $R_g$ is the guiding radius of the orbit.

An induced error in $z$ of $\Delta z$ will cause $\Delta A_z \sim \Delta z$.
Similarly, an induced error in $R$ will cause $\Delta A_R \sim \Delta R$.
Making the assumption that $\kappa$ and $\nu$ are constant, we have that
\beq\label{eq:actions_epi_error}
\begin{split}
\Delta J_{\phi} = \Delta R v_c &\implies \frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R} \\
\Delta J_R = 2 \kappa \Delta A_R &\implies \frac{\Delta J_R}{J_R} = \frac{2\Delta R}{A_R} \\
\Delta J_z = 2 \nu \Delta A_z &\implies \frac{\Delta J_z}{J_z} = \frac{2\Delta z}{A_z}\text{.}
\end{split}
\eeq
Therefore, errors in each action are linearly proportional to the errors in
either the midplane or cylindrical radius. While in detail $\kappa$ and $\nu$
are not constant, we will find that this order of magnitude estimate matches
the numerical estimates quite well.

There is a slight complication in the formula for computing $\Delta J_R/J_R$
and $\Delta J_{\phi}/J_{\phi}$. In a moment we will consider the effect of an inaccuracy
in the distance from the Sun to the Galactic center on actions. However, this
inaccuracy induces an error in $x$, or $\Delta x$. This is not the same as an
inaccuracy in $\Delta R$, but they are related by the following argument. An
error $\Delta x$ will induce an erroneous radius $R_{\text{err}}$ related by
the formula,
\beq
(x+\Delta x)^2 + y^2 = R_{\text{err}}^2\text{.}
\eeq
Keeping only terms to first order in $\Delta x$, we have that,
\beq
\begin{split}
R_{\text{err}}^2 &= R^2 - 2 R \cos{\theta} \Delta x \\
\implies \Delta R &\equiv \abs{R_{\text{err}} - R} = \abs{\cos{\theta}} \Delta x\text{.}
\end{split}
\eeq
Averaging over the circle, we therefore have that,
\beq
\avg{\Delta R} = \frac{2}{\pi} \Delta x\text{.}
\eeq
Therefore, the error in $J_R$ and $J_{\phi}$ are related to $\Delta x$ by,
\beq\label{eq:Jr_epi_dx}
\begin{split}
\frac{\Delta J_R}{J_R} &= \frac{4 \Delta x}{\pi A_R} \\
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{2 \Delta x}{\pi R}\text{.}
\end{split}
\eeq

A similarly rough argument can be made to arrive at the following formulae for
the impact of velocity offsets on the actions,
\beq\label{eq:actions_epi_error}
\begin{split}
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{\Delta v_{\phi}}{v_c} \\
\frac{\Delta J_R}{J_R} &= \frac{\Delta v_R}{v_{R,\text{max}}} \\
\frac{\Delta J_z}{J_z} &= \frac{\Delta v_z}{v_{z,\text{max}}}\text{,}
\end{split}
\eeq
where $v_{R,\text{max}}$ and $v_{z,\text{max}}$ are the maximum radial and
vertical velocities of the orbit, respectively.

>>>>

\subsection{Epicyclic Approximation} \label{ssec:epi_action}
Before turning to numerical methods, we first write down analytically the systematic error in the actions induced from either a midplane offset or a Galactocentric radius
offset. We use the epicyclic approximation, which assumes that the motion in
the $z$ and $R$ components of the orbit are decoupled and follow simple
harmonic motion \citep[][section 3.2 and references therein]{2008gady.book.....B}. This approximation is an excellent description of
thin disk-like orbits, and a good description of thick disk-like orbits, in an axisymmetric potential (i.e. ignoring the influence of the Galactic bar and spiral arms). We
also make the assumption of a perfectly flat circular velocity curve with
velocity $v_c$ \citep[decent near the Solar circle, e.g.][]{2017MNRAS.465...76M}.

Under this approximation, we can write down the vertical and radial components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
z(t) &= A_z \sin{(\kappa t + \alpha)}\\
R(t) &= A_R \sin{(\nu t + \beta)}\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the vertical and radial (or epicyclic)
frequencies, $A_z$ and $A_R$ are the amplitudes of the stellar motion in the
vertical and radial coordinates, and $\alpha$ and $\beta$ are arbitrary
constants. Similarly, the velocities of the orbit are given by
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_z(t) &= A_z \kappa \cos{(\kappa t + \alpha)}\\
v_R(t) &= A_R \nu \cos{(\nu t + \beta)}\text{.}
\end{split}
\eeq

We can write down the actions in this approximation as
\citep[][Section~3.5.3b]{2008gady.book.....B},
\beq\label{eq:actions_epi}
\begin{split}
J_{\phi} &= R_g v_c \\
J_R &= \kappa A_R^2 \\
J_z &= \nu A_z^2\text{,}
\end{split}
\eeq
where $R_g$ is the guiding radius of the orbit.

An offset $\Delta z$ will induce $\Delta A_z \sim \Delta z$.
Similarly, an offset $\Delta R$ will induce $\Delta A_R \sim \Delta R$.
Assuming $\kappa$ and $\nu$ are constant, 
\beq\label{eq:actions_epi_error}
\begin{split}
\Delta J_{\phi} = \Delta R v_c &\implies \frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R} \\
\Delta J_R = 2 \kappa \Delta A_R &\implies \frac{\Delta J_R}{J_R} = \frac{2\Delta R}{A_R} \\
\Delta J_z = 2 \nu \Delta A_z &\implies \frac{\Delta J_z}{J_z} = \frac{2\Delta z}{A_z}\text{.}
\end{split}
\eeq
Therefore, systematic errors in each action are linearly proportional to offsets in
either the midplane ($z$ coordinate) or cylindrical radius ($R$ coordinate). While in general $\kappa$ and $\nu$ are not constant, we find that this order of magnitude estimate matches our numerical calculations quite well.

%There is a slight complication in the formula for computing $\Delta J_R/J_R$
%and $\Delta J_{\phi}/J_{\phi}$. 
We will later consider the effect on actions of an inaccuracy in the distance from the Sun to the Galactic center, which introduces an offset in the $x$ coordinate, $\Delta x$, of each star when converting to a Galactocentric coordinate system. An offset $\Delta x$ results in an erroneous radius $R_{\text{err}}$ related by
the formula,
\beq
(x+\Delta x)^2 + y^2 = R_{\text{err}}^2\text{.}
\eeq
Keeping only terms to first order in $\Delta x$, we have that,
\beq
\begin{split}
R_{\text{err}}^2 &= R^2 - 2 R \cos{\theta} \Delta x \\
\implies \Delta R &\equiv \abs{R_{\text{err}} - R} = \abs{\cos{\theta}} \Delta x\text{.}
\end{split}
\eeq
Averaging over the circle, we therefore have that,
\beq
\avg{\Delta R} = \frac{2}{\pi} \Delta x\text{.}
\eeq
Therefore, the error in $J_R$ and $J_{\phi}$ are related to $\Delta x$ by,
\beq\label{eq:Jr_epi_dx}
\begin{split}
\frac{\Delta J_R}{J_R} &= \frac{4 \Delta x}{\pi A_R} \\
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{2 \Delta x}{\pi R}\text{.}
\end{split}
\eeq

A similar derivation results in formulae for
the impact of velocity offsets on the actions:
\beq\label{eq:actions_epi_error}
\begin{split}
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{\Delta v_{\phi}}{v_c} \\
\frac{\Delta J_R}{J_R} &= \frac{\Delta v_R}{v_{R,\text{max}}} \\
\frac{\Delta J_z}{J_z} &= \frac{\Delta v_z}{v_{z,\text{max}}}\text{,}
\end{split}
\eeq
where $v_{R,\text{max}}$ and $v_{z,\text{max}}$ are the maximum radial and
vertical velocities of the orbit, respectively.
